Evolutionary Learning of Policies for MCTS Simulations

### 2012-03-18 thesis/paper/paper.tex
Working through revisions suggested by reviewers.

\documentclass{acm_proc_article-sp}
\usepackage[latin1]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{url}
\usepackage{havannah}
\newcommand{\black}{\texttt{Black}}
\newcommand{\white}{\texttt{White}}
\newcommand{\blacks}{\texttt{Black's}}
\newcommand{\whites}{\texttt{White's}}
\newcommand{\loc}[1]{\texttt{#1}}

\title{Learning to Bias MCTS Playouts}
\numberofauthors{2}
\author{ 
% You can go ahead and credit any number of authors here,
% e.g. one 'row of three' or two rows (consisting of one row of three
% and a second row of one, two or three).
%
% The command \alignauthor (no curly braces needed) should
% precede each author name, affiliation/snail-mail address and
% e-mail address. Additionally, tag each line of
% affiliation/address with \affaddr, and tag the
% e-mail address with \email.
%
% 1st. author
\alignauthor
James Pettit \\
       \affaddr{U.C. Santa Cruz}\\
       \affaddr{Santa Cruz, California}\\
       \email{ james.l.pettit@gmail.com }
% 2nd. author
\alignauthor
David Helmbold \\
       \affaddr{U.C. Santa Cruz}\\
       \affaddr{Santa Cruz, California}\\
       \email{dph@soe.ucsc.edu}
}

@@

\begin{abstract}
Monte-Carlo Tree Search (MCTS) 
grows a partial game tree and uses a large number of random simulations to
approximate the values of the nodes.   
It has proven effective in games with such as Go and Hex where the large search space and
difficulty of evaluating positions cause difficulties for standard methods.
The best MCTS players use carefully hand-crafted rules to bias the random simulations.
Obtaining good  hand-crafting rules is a very difficult process, as even rules promoting better
simulation play can result in a weaker MCTS system~\cite{gelly2006modification}.
Our Hivemind system uses evolutionary strategies to automatically learn effective rules for biasing the 
random simulations.
We have built a MCTS player using Hivemind for the game Hex. 
The Hivemind learned rules result in a 90\% win rate against a baseline MCTS system, and 
significant improvement against the computer Hex world champion, MoHex.
\end{abstract}

\section{Introduction}
Abstract strategy games provide an attractive testbed for artificial intelligence and machine learning applications. The game framework provides immediate feedback on algorithm performance. The choice of game and rules can also be experimentally varied to give a range of difficulty and computational complexity levels. Starting with mathematical game theory research, the most common technique for game AI has been brute force search of the tree of legal moves. This technique, highly tuned and optimized, is what led to Chinook's ascendance to World Champion in 1994 and Deep Blue's eventual victory in 1997 over then World Champion Gary Kasparov. These victories cemented brute force tree search as standard techniques in abstract strategy games, to the point where computers constructing endgame tables have been strong enough to find sequences lasting for hundreds of moves. Kasparov himself suggested a version of Chess that allowed the player to consult a computer for move advice. Chinook went so far as to solve checkers, proving that perfect play will end in a draw. 
Several games, such as  Checkers, Chess, and Othello, have good enough evaluation functions and few enough legal moves so that that brute force tree search techniques can be effective.

In contrast, cutting and connecting games like Go, Hex, and Havannah, have much larger search trees and long-range interactions make it difficult to create accurate evaluation functions.
In these domains, Monte-Carlo Tree Search (MCTS) has proven effective~\cite{}, 
with the world champion Go and Hex players both using MCTS~\cite{}.
MCTS picks its next move by growing a partial game tree while using
a large number of randomized simulations to evaluate the possible moves.
Each simulation first uses a child-selection policy to navigate through the partial game tree.
After reaching a leaf of the partial game tree, a randomized continuation policy is used to complete the game 
and determine a winner.  
Finally, an update policy is used to update the partial game tree with the result of the simulation.
Although all three policies have an effect on the algorithm's playing strength, here we focus on
continuation policies for the game Hex.
Section~\ref{mcts} contains a fuller description of MCTS.

MoGo specifically, and Go programs in general, have the benefit of a rich history of play. Thousands of expert games are available for analysis. Some approaches to improving the simulation policy have focused on using expert games in training weights to direct the playout~\cite{chaslot2010adding}. More generally, common patterns are used in a hand-crafted policy to select ``urgent'' moves to play in a local area around the last played move. Hex does have one simple pattern in the form of a bridge~\cite{anshelevich2002hierarchical}, but few easy paths to crafting a better continuation policies.

The Hivemind approach automatically learns better continuation policies from self-play. 
The usual process of improvement requires hand-tuned weights that bias the continuation policy. 
Any change must be tested to detect the change in strength it gives,
and different weight parameters can have synergistic effects.
Small changes in overall algorithm strength require thousands of games of testing to detect. 
Often the improvement made by a small change in weight can be so small it will go unnoticed. 
Automatic weight tuning can improve this process, but still requires thousands of games to test. 
Recent work in this area has focused on learning weights for a small number of patterns using expert knowledge and reinforcement learning~\cite{silver2009monte}. 
Hivemind is an alternative approach using evolutionary learning. 
Because it relies on self-play, Hivemind avoids the need for extensive expert knowledge or a strong outside player.
Although our experiments have focused on Hex, 
the Hivemind framework is suitable for generic Monte-Carlo Tree Search, and has been implemented for
both Go and Hex.

@@

The evolutionary strategies approach used by Hivemind is surprisingly stable.
In 10 different evolutionary runs, the win rates over the unbiased continuation policy remained
between 87\% and 93.5\%.  
Furthermore, the learned strategies generalize well to different board sizes.
Despite using $7 \times 7$ boards for faster training, 
the learned biases performed as well (and usually better!) against the simple policies on $11 \times 11$ and $13 \times 13$ boards.

@@

\section{Background: The Game of Hex}
\label{s:hex}
Hex is a two player, full-information, abstract strategy game played on an $N$x$N$ grid of hexagons. The players are denoted by  a color, either \black{} and \white. 
The players alternate placing markers of their color on empty hexagons. 
The object of the game is to form an unbroken chain from one side of the board to the other. 
Figure \ref{fig:13x13inprogress} shows an example game in progress on a 13x13 board, note that the numbers indicate
the order in which the markers were played and have no effect on the game.
Figure \ref{fig:13x13finished} shows the final position, \black{} having just played the winning move (71), completing an unbroken chain connecting both his sides.

Hex is a cutting and connecting game that
has properties that make it attractive for AI research. 
Once placed, a piece never leaves the board. 
The only criterion for a legal move is that the hexagon be unoccupied (empty). 
Gale '79 proved the game cannot end in a draw, simplifying minimax methods. 
Nash, one of the inventors of the game, used a strategy stealing argument to prove the existence of a winning strategy for the first player, but the proof is by reduction and offers no insight into the winning strategy.
Solutions for all opening positions for 7x7 and 8x8 boards have been found~\cite{henderson2009solving}. 
A good portion, but not all of the 9x9 openings have also been solved. 
Nash proposed 11x11 as the ideal board size, but 13x13 has been more common in competitions, especially as the ability of computer players improves.

There are three exceptionally strong Hex programs.  Gabor Melis' \emph{Six} is the previous world champion (up to 2007). 
Six uses traditional alpha-beta search with an evaluation function based on an electrical circuit model (proposed by Anshelevich 2002)~\cite{anshelevich2002hierarchical}. 
\emph{Wolve}, from the University Alberta, also uses the Anshelevich algorithm, but more aggressively prunes provably inferior moves. 
\emph{MoHex}, also from Alberta, is based on the newer Monte-Carlo Tree Search technique, leveraging the open source \emph{Fuego} library. 
MoHex is the current world-champion, having finished ahead of Wolve in 2008, 2009 and 2010.

Go is a much older game than Hex or Havannah, and much research has been done on Go playing algorithms. 
Despite its simpler rules, Hex has a similar branching factor and position evaluation complexity as Go. 
Although we emphasize Hex here, the techniques discussed generalize to any large branching game played on a regular board
(such as Go or Havannah).

\begin{figure}[tb]
\resizebox{3.5in}{!}{	\begin{HexBoard}[board size=13]
		\HGame{c4,f6,e11,i8,h8,i1,i6,d5,i3,h9,h5,h4,i4,g9,f9,g8,f7,f8,e8}
	\end{HexBoard}
	}
	\caption{13x13 Hex Game, \white{} to play.}
	\label{fig:13x13inprogress}
\end{figure}

\begin{figure}[tb]
\resizebox{3.5in}{!}{	\begin{HexBoard}[board size=13]
		\HGame{c4,f6,e11,i8,h8,i1,i6,d5,i3,
				   h9,h5,h4,i4,g9,f9,g8,f7,f8,
				   e8,e9,c10,k8,c9,g6,g7,c11,
				   d10,h6,h7,i5,j5,j4,k4,k3,
				   m2,l3,m3,l4,c6,e11,l5,m4,
				   d11,b7,e6,d12,c12,c7,e7,f5,
				   e5,d9,d8,f4,e4,f3,a2,e3,d6,
				   c3,d3,d4,b5,e2,d2,d1,e1,b13,
				   c13,c5,b6}
	\end{HexBoard}
	}
	\caption{Finished 13x13 Hex Game, \black{}'s play 71 (at \loc{b6}) gives him an unbroken path from top to bottom.}
	\label{fig:13x13finished}
\end{figure}

@@

\begin{figure}
	\begin{center}
%	\centerline{\begin{HexBoard}[board size=5]
%		\HGame{a3,c2}
%	\end{HexBoard}
%	}
	\includegraphics[width=3.0in]{graphics/tree.png}
	\end{center}
	\caption{Partial MC Search Tree}
	\label{fig:tree}
\end{figure}

@@

The basic Monte-Carlo technique for Computer Go was proposed by Bru\"{u}gmann~\cite{brugmann1993monte}. His program \emph{Mango} effectively used a ``flat'' search tree (1-ply) and simply ran simulations from each candidate move to estimate its value. To contrast, MCTS uses Monte-Carlo simulations to grow a partial  search tree. 
The UCT algorithm was developed alongside MCTS to direct the growth of the search tree~\cite{gelly2006exploration}. 
At any point, the search can be stopped and the current state of the tree used to select a move to play.

All nodes in the tree correspond to a fixed board position, i.e. the occupancy state of the board and which player is next to move. 
Nodes also have statistics associated with their position:  the number of winning simulations performed that used the node, and the total number of simulations that included the node. 
For each simulation the algorithm will start at the root node, recursively traverse the (partial) tree until it reaches a leaf node.
The simulation continues from the position represented by this leaf node until a winner is determined.
The result of the simulation used to update the statistics at nodes in the tree.
If the leaf node has been visited enough times (50 in our case), then it is expanded and nodes for all possible moves from 
that position are added to the search tree.

The effectiveness of MCTS depends on three policies:
\begin{itemize}
\item{The \emph{child-selection} policy recursively traverses the tree, selecting promising moves for further exploration.}
\item{The \emph{continuation} policy randomly or semi-randomly ``plays out'' the game from a leaf position in the tree until a final position.}
\item{The \emph{update} policy modifies the win/loss statistics of nodes based on the simulation results, and determines when to grow the search tree.}
\end{itemize}
Figure \ref{fig:tree} shows a partial Monte-Carlo search tree for the given position. In the example, the child-selection policy visited the node for \texttt{Black E4}, then \texttt{White A1}. 
The continuation policy ran, completing the game from the positiion with \texttt{E4} occupied by \black\ and \texttt{A1} occupied by \white.
This resulted in a loss for \black. 
The update policy choose to update all nodes along the visited path, incrementing the wins for \white\ moves and incrementing the losses for \black\ moves.

\begin{lstlisting}[float,frame=single,language=Python,caption=MCTS Algorithm Pseudocode]
def search(root):
  for i in range(simulations):
    step(root)

def play_from(node) returns outcome:
  if node.visits < threshold:
    outcome = simulate(node)
  else:
    child = select(node)
    outcome = play_from(child)
  if outcome is win for node.player:
    node.wins++
  node.visits++
  return outcome
\end{lstlisting}

\subsection{Child Selection Policy}
The \texttt{select} function defines the child selection policy. It must return the child of the given node to step into next. 
In general, child selection must trade off between \emph{exploitation} and \emph{exploration}. 
It is important to ``exploit" those child nodes with high win-rates, both
to discover possible weaknesses that would result in the opponent winning and to obtain accurate win-rate estimates. 
Simultaneously, a child with a low number of visits must also be explored since a low win-rate
could be due to a few unlucky simulations.
This exploration/exploitation tradeoff is related to a variance reduction problem.  
We want very low variance on the estimates for the good moves in order to maximize the chances of making the best move.
On the other hand, if the variance of a seemingly poor move is very high, then there is a significant chance that that move
could actually be very good.

@@

The \texttt{continuation} function plays out the rest of the game from the current position. 
The simplest policy, which we call the \emph{default policy},
 plays uniformly at random from the legal moves until the game ends (in Hex, when one player has a winning path). 
The continuation policy has a large effect on the strength of the MCTS player.
% and the Go literature has a large body of work dedicated to improving it.
Initial research in MCTS Go players focused on using a strong continuation policy, with expert-derived heuristics to select the next move~\cite{chaslot2010adding}. 
The top Computer Go programs all have complex hand-crafted policies with rules to examine the current position and select a ``good'' move to play, possibly with a stochastic choice between several candidates. 
Only when none of the rules match do they fall back to a uniform random move. 
It was quickly found that building a stronger continuation policy does not guarantee better performance when used in the full tree search~\cite{gelly2006modification}. 
In other words, when the continuation policies are used as players, the relative strengths of the continuation policies are poor predictors
of how well the MCTS algorithms relying on these continuation policies will perform.
This surprising result may be due to stronger continuation policies under-exploring the search space.
In any case, this means that hand crafting an effective continuation policy requires benchmarking and frequent tests of the entire
MCTS system as small changes to the simulation policy can have unpredictable effects on overall performance.
This illustrates the need for automated systems for tuning the continuation policy.

@@

\subsection{Update Policy}
Once a simulation is finished, the result is used to update the search tree.
Basic MCTS systems updated the win and visit statistics only along the traversed path.
It was quickly discovered that more aggressive update strategies are often profitable.
One simple example of this is the All-Moves-As-First (AMAF) heuristic~\cite{}
that exploits the fact that moves played in a simulation could have been played in any order. 
Thus, a move deep in the simulation could be fictitiously treated as the ``first'' move from an intermediate position. 
In addition to updating the statistics of the nodes in the path, all children of nodes on the path corresponding to moves made later
in the simulation have their AMAF win and visit counts updated. 
Variations on AMAF have been tested, including Permutations AMAF, which updates a node if any permutation of the simulation contained the node's position, and Some-Moves-As-First, which attempts to restrict the number of siblings updated.

@@

% TODO
\textbf{does MoHex do something fancier?}
% MoHex uses one local pattern - bridge preservation, and AMAF updates with RAVE (http://webdocs.cs.ualberta.ca/~hayward/hex/#mohex)

@@

\subsection{Biasing the Continuation -- Incorporate into earlier subsections}\label{bias}

Gelly and others have shown that using heuristics in both the expansion and playout policy can prove extremely beneficial~\cite{gelly2006modification,gelly2008achieving}. However, this can require expert knowledge (heuristics) that is hard to obtain or not available for some games (particularly new games). They also found found that improving the performance of playout policy did not guarantee an improvement in the tree search. For example, the playout policy can implicitly prune good moves, never allowing them to be played. In Go, the policy must correctly handle complicated situations like nakade (life and death). If it fails to correctly play out a complex position, the tree might fail to see an opponent response. Much time and energy has been spent on hand-tuning the policies of the top engines to correctly handle complicated tactical situations. If there is some small probability of playing a purely random move, the convergence properties of the UCT algorithm will still hold, but can require a prohibitive number of simulations.
In fact, one of the top Go engines, MoGo, uses a near-zero exploration coefficient in its child selection policy~\cite{gelly2007combining}. This puts most of the control of the search in the hands of the simulation policy.
%
%\section{Problem Statement  -- delete? Incorporate points into MCTS section?}\label{problem}
%MoGo specifically, and Go programs in general, have the benefit of a rich history of play. Thousands of expert games are available for analysis. Some approaches to improving the simulation policy have focused on using expert games in training weights to direct the playout~\cite{chaslot2010adding}. More generally, common patterns are used in a hand-crafted policy to select ``urgent'' moves to play in a local area around the last played move. Hex does have one simple pattern in the form of a bridge~\cite{anshelevich2002hierarchical}, but few easy paths to crafting a better simulation policy.
%
%It would be ideal to automatically learn a better playout policy from self-play. The usual process of improvement requires hand-tuned weights that modify the playout policy. Any change must be tested to detect the change in strength it gives. Problematically, small changes in overall algorithm strength require thousands of games of testing to detect. Many times the improvement made by a small change in weight can be so small it will go unnoticed. Automatic weight tuning can improve this process, but still requires thousands of games to test. Recent work in this area has focused on learning weights for a small number of patterns using expert knowledge and reinforcement learning~\cite{silver2009monte}. This thesis presents a slightly different approach using evolutionary learning. It solves the problem of requiring a strong outside player or expert knowledge by purely using games of self-play.
%

@@

Hivemind is a system that learns to play abstract strategy games on a regular board. 
The system consists of three modules: the evolutionary learning process, a Monte-Carlo based UCT search tree implementation, and a ``Game Tracker interface'' used by the tree search to maintain game state information. 
The design was influenced by Fuego, although Hivemind is written in a different language (Google's Go~\cite{golang}). 
In particular, the tracker interface is very similar to Fuego's subclassing system for implementing multiple games~\cite{Fuego}. 
Hivemind's tracker interface is currently implemented for the games of Hex, Go, and Tic-Tac-Toe. 
The major contribution of Hivemind is the integration of evolutionary learning to automatically learn appropriate continuation policies.
Hivemind is also open-source software, available at \url{http://github.com/etherealmachine/hivemind}.
Preliminary work tried learning with particle swarm optimization, but we found that evolutionary strategies gave
superior results.

Using evolutionary strategies to learn continuation policies requires that the continuation policies are encoded as vectors.
We first describe this encoding.  Section~\ref{s:tracker} describes the Game Tracker module, and Section~\ref{s:es} describes the
use of Evolutionary Strategies.

@@

\begin{figure}
	\begin{center}
	\includegraphics[width=3.0in]{graphics/local-pattern.png}
	\caption{A 4x4 board with the 3 patterns shown around \blacks\ last move. Example weights are shown in the center vertex.}
	\label{fig:localpattern}
	\end{center}
\end{figure}

>>>>

\documentclass{acm_proc_article-sp}
\usepackage[latin1]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{url}
\usepackage{havannah}
\newcommand{\hblack}{\texttt{Black}}
\newcommand{\hwhite}{\texttt{White}}
\newcommand{\hblacks}{\texttt{Black's}}
\newcommand{\hwhites}{\texttt{White's}}
\newcommand{\loc}[1]{\texttt{#1}}

\title{Evolutionary Learning of Policies for MCTS Simulations}
\numberofauthors{2}
\author{ 
% You can go ahead and credit any number of authors here,
% e.g. one 'row of three' or two rows (consisting of one row of three
% and a second row of one, two or three).
%
% The command \alignauthor (no curly braces needed) should
% precede each author name, affiliation/snail-mail address and
% e-mail address. Additionally, tag each line of
% affiliation/address with \affaddr, and tag the
% e-mail address with \email.
%
% 1st. author
\alignauthor
James Pettit \\
       \affaddr{U.C. Santa Cruz}\\
       \affaddr{Santa Cruz, California}\\
       \email{ jpettit@soe.ucsc.edu}
% 2nd. author
\alignauthor
David Helmbold \\
       \affaddr{U.C. Santa Cruz}\\
       \affaddr{Santa Cruz, California}\\
       \email{dph@soe.ucsc.edu}
}

@@

\begin{abstract}
Monte-Carlo Tree Search (MCTS) 
grows a partial game tree and uses a large number of random simulations to
approximate the values of the nodes.   
It has proven effective in games with such as Go and Hex where the large search space and
difficulty of evaluating positions cause difficulties for standard methods.
The best MCTS players use carefully hand-crafted rules to bias the random simulations.
Obtaining good  hand-crafting rules is a very difficult process, as even rules promoting better
simulation play can result in a weaker MCTS system~\cite{gelly2006modification}.
Our Hivemind system uses evolution strategies to automatically learn effective rules for biasing the 
random simulations.
We have built a MCTS player using Hivemind for the game Hex. 
The Hivemind learned rules result in a 90\% win rate against a baseline MCTS system, and 
significant improvement against the computer Hex world champion, MoHex.
\end{abstract}

\keywords{Monte Carlo Tree Search, Computer Hex, Evolution Strategies, Evolutionary Learning, Adversarial Search, Monte Carlo}

\section{Introduction}
Abstract strategy games provide an attractive testbed for artificial intelligence and machine learning applications. The game framework provides immediate feedback on algorithm performance. 
%The choice of game and rules can also be experimentally varied to give a range of difficulty and computational complexity levels. 
Starting with mathematical game theory research, the most common technique for game AI has been brute force search of the tree of legal moves. 
This technique, highly tuned and optimized, is what led to Chinook's ascendance to World Champion in 1994 and Deep Blue's eventual victory in 1997 over then World Champion Gary Kasparov. These victories cemented brute force tree search as standard techniques in abstract strategy games, to the point where computers constructing endgame tables have been strong enough to find sequences lasting for hundreds of moves. Kasparov himself suggested a version of Chess that allowed the player to consult a computer for move advice. Chinook went so far as to solve checkers, proving that perfect play will end in a draw. 
Several games, such as  Checkers, Chess, and Othello, have good enough evaluation functions and few enough legal moves so that that brute force tree search techniques can beat the best human players.

In contrast, cutting and connecting games (like Go, Hex, and Havannah)
have much larger search trees and long-range interactions make it difficult to create accurate evaluation functions.
In these domains, Monte-Carlo Tree Search (MCTS) has proven effective, 
with the computer world champion Go and Hex players both using MCTS.
MCTS picks its next move by growing a partial game tree while using
a large number of randomized simulations to evaluate the possible moves.
Each simulation first uses a child-selection policy to navigate through the partial game tree.
After reaching a leaf of the partial game tree, a randomized continuation policy is used to play out the rest of the game 
so a winner can be determined.
Finally, an update policy is used to update the partial game tree with the result of the simulation.
Although all three policies have an effect on the algorithm's playing strength, here we focus on
continuation policies for the game Hex.
Section~\ref{mcts} contains a fuller description of MCTS.

Go programs have the benefit of a rich history of play. Thousands of Go expert games are available for analysis. Some approaches to improving the simulation policy have focused on using expert games in training weights to direct the playout~\cite{chaslot2010adding}. More generally, common patterns are used in a hand-crafted policy to select ``urgent'' moves to play in a local area around the last played move. Hex does have one simple pattern in the form of a bridge~\cite{anshelevich2002hierarchical}, but few easy paths to crafting a better continuation policies.

The Hivemind approach automatically learns better continuation policies from self-play. 
The usual process of improvement requires hand-tuned weights that bias the continuation policy. 
Any change must be tested to detect the change in strength it gives,
and different weight parameters can have synergistic effects.
Detecting small changes in overall algorithm strength require thousands of testing games on a full-sized board, and
the improvement made by a small change in weight can go unnoticed.
Automatic weight tuning can improve this process, but still requires thousands of learning games. 
Recent work in this area has focused on learning weights for a small number of patterns using expert knowledge and reinforcement learning~\cite{silver2009monte}. 
Hivemind is an alternative approach using evolutionary learning.
Because it relies on self-play, Hivemind avoids the need for extensive expert knowledge or a strong outside player.
By using an evolutionary approach, Hivemind can handle the subtle interactions between weights that can have a strong effect on overall playing strength.
Although our experiments have focused on Hex, 
the Hivemind framework is suitable for generic Monte-Carlo Tree Search, and has been implemented for
both Go and Hex.

Our experiments answer several questions, including: \\
$\bullet$  How beneficial are the learned policies ?  \\
$\bullet$   How stable is the evolutionary learning process? \\
$\bullet$   Do the learned policies work on different board sizes?

@@

The evolution strategies approach used by Hivemind is surprisingly stable.
In 10 different evolutionary runs, the win rates over the unbiased continuation policy remained
between 87\% and 93.5\%.  
Furthermore, the learned strategies generalize well to different board sizes.
Despite using $7 \times 7$ boards for faster training, 
the learned biases performed as well (and usually better!) against the simple policies on $11 \times 11$ and $13 \times 13$ boards.

@@

\section{Background: The Game of Hex}
\label{s:hex}
Hex is a two player, full-information, abstract strategy game played on an $N$x$N$ grid of hexagons. The players are \hblack{} and \hwhite{} and alternate placing markers of their color on empty hexagons. 
The object of the game is to form an unbroken chain from one side of the board to the other. 
Figure \ref{fig:13x13inprogress} shows an example game in progress on a 13x13 board, note that the numbers indicate
the order in which the markers were played and have no effect on the game.
Figure \ref{fig:13x13finished} shows the final position, \hblack{} having just played the winning move (71), completing an unbroken chain connecting both his sides.

Hex is a cutting and connecting game whose high computational complexity and simple scoring mechanism makes it attractive for AI research. 
Once placed, a piece never leaves the board. 
The only criterion for a legal move is that the hexagon be unoccupied (empty). 
Gale '79 proved the game cannot end in a draw, simplifying minimax methods. 
Nash, one of the inventors of the game, used a strategy stealing argument to prove the existence of a winning strategy for the first player, but the proof is by reduction and offers no insight into the winning strategy.
Solutions for all opening moves on a 7x7 and 8x8 board have been found~\cite{henderson2009solving}. 
Of the 81 possible 9x9 openings, only 13 remain unsolved (taking symmetry into account). 13x13 is the default board size in computer Hex competitions.

There are three Hex programs that have historically competed for world champion status in the Computer Olympiad. Gabor Melis' \emph{Six} is the previous world champion (up to 2007). 
Six uses traditional alpha-beta search with an evaluation function based on an electrical circuit model (proposed by Anshelevich 2002)~\cite{anshelevich2002hierarchical}. 
\emph{Wolve}, from the University Alberta, also uses the Anshelevich algorithm, but more aggressively prunes provably inferior moves. 
\emph{MoHex}, also from Alberta, is based on the newer Monte-Carlo Tree Search technique, leveraging the open source \emph{Fuego} library. 
MoHex is the current world-champion, having finished ahead of Wolve in 2008, 2009 and 2010.

%Go is a much older game than Hex, and much research has been done on Go playing algorithms. 
Despite its simpler rules, Hex has a similar branching factor as Go and long-range interactions make it difficult to 
evaluate Hex positions.
Although we emphasize Hex here, the techniques we propose generalize to any cutting and connecting game
(such as Go or Havannah).

\begin{figure}[tb]
\resizebox{3.5in}{!}{	\begin{HexBoard}[board size=13]
		\HGame{c4,f6,e11,i8,h8,i1,i6,d5,i3,h9,h5,h4,i4,g9,f9,g8,f7,f8,e8}
	\end{HexBoard}
	}
	\caption{13x13 Hex Game, \hwhite{} to play.}
	\label{fig:13x13inprogress}
\end{figure}

\begin{figure}[tb]
\resizebox{3.5in}{!}{	\begin{HexBoard}[board size=13]
		\HGame{c4,f6,e11,i8,h8,i1,i6,d5,i3,
				   h9,h5,h4,i4,g9,f9,g8,f7,f8,
				   e8,e9,c10,k8,c9,g6,g7,c11,
				   d10,h6,h7,i5,j5,j4,k4,k3,
				   m2,l3,m3,l4,c6,e11,l5,m4,
				   d11,b7,e6,d12,c12,c7,e7,f5,
				   e5,d9,d8,f4,e4,f3,a2,e3,d6,
				   c3,d3,d4,b5,e2,d2,d1,e1,b13,
				   c13,c5,b6}
	\end{HexBoard}
	}
	\caption{Finished 13x13 Hex Game, \hblack{}'s play 71 (at \loc{b6}) gives him an unbroken path from top to bottom.}
	\label{fig:13x13finished}
\end{figure}

@@

\begin{figure}
	\begin{center}
%	\centerline{\begin{HexBoard}[board size=5]
%		\HGame{a3,c2}
%	\end{HexBoard}
%	}
	\includegraphics[width=3.0in]{graphics/tree.pdf}
	\end{center}
	\caption{Partial MC Search Tree}
	\label{fig:tree}
\end{figure}

@@

The basic Monte-Carlo technique for Computer Go was proposed by Br\"{u}gmann~\cite{brugmann1993monte}. His program \emph{Mango} effectively used a ``flat'' search tree (1-ply) and simply ran simulations from each candidate move to estimate its value. 
Later MCTS techniques use Monte-Carlo simulations to grow a partial  search tree. 
The UCT algorithm was developed alongside MCTS to direct the growth of the search tree~\cite{gelly2006exploration}. 
At any point, the search can be stopped and the current state of the tree used to select a move to play.

All nodes in the tree correspond to a fixed board position, i.e.~the occupancy state of the board and which player is next to move. 
Nodes also have statistics associated with their position:  the number of winning simulations performed that used the node, and the total number of simulations that included the node. 
For each simulation the algorithm will start at the root node, recursively traverse the (partial) tree until it reaches a leaf node.
The simulation continues from the position represented by this leaf node until a winner is determined.
The result of the simulation is used to update the statistics at nodes in the tree
(see Listing 1).
If the leaf node has been visited enough times (50 in our case), then it is expanded and nodes for all possible moves from 
that position are added to the search tree.

The effectiveness of MCTS depends on three policies:
\begin{itemize}
\item{The \emph{child-selection} policy recursively traverses the tree, selecting promising moves for further exploration.}
\item{The \emph{continuation} policy randomly or semi-randomly ``plays out'' the game from a leaf position in the tree until a final position.}
\item{The \emph{update} policy modifies the win/loss statistics of nodes based on the simulation results, and determines when to grow the search tree.}
\end{itemize}

Gelly and others have shown that using heuristics in both the child-selection and continuation policies 
can prove extremely beneficial~\cite{gelly2006modification,gelly2008achieving} for Go.
However, this requires expert knowledge (heuristics) that is hard to obtain or not available for some games (particularly new games).

Figure \ref{fig:tree} shows a partial Monte-Carlo search tree for the given position. In the example, the child-selection policy visited the node for \texttt{Black E4}, then \texttt{White A1}. 
The continuation policy ran, completing the game from the positiion with \texttt{E4} occupied by \hblack\ and \texttt{A1} occupied by \hwhite.
This resulted in a loss for \hblack. 
The update policy choose to update all nodes along the visited path, incrementing the wins for \hwhite\ moves and incrementing the losses for \hblack\ moves.

\begin{lstlisting}[float,frame=single,language=Pascal,caption=MCTS Algorithm Pseudocode]
function search(root):
  for i in range(simulations):
    simulate_from (root)

function simulate_from(node) returns outcome
  if node is leaf
    outcome =  continue_from (node)
  else:
    child = select_child (node)
    outcome = simulate_from (child)
  node.visits++
  if outcome is  node.player win
    node.wins++
  if (node.visits ==  50) expand (node)
  return outcome
\end{lstlisting}

\subsection{Child Selection Policy}
The \verb+select_child+ function defines the child selection policy. It must return the child of the given node to step into next. 
In general, child selection must trade off between \emph{exploitation} and \emph{exploration}. 
It is important to ``exploit" those child nodes with high win-rates, both
to discover possible weaknesses and to obtain highly accurate win-rate estimates. 
Simultaneously, a child with a low number of visits must also be explored since a low win-rate
could be due to a few unlucky simulations.
This exploration/exploitation tradeoff is related to a variance reduction problem.  
We want very low variance on the estimates for the good moves in order to maximize the chances of making the best move.
On the other hand, if the variance of a seemingly poor move is very high, then there is a significant chance that that move
could actually be very good.

@@

The \verb+continue_from+ function plays out the rest of the game from the current position. 
The simplest policy, which we call the \emph{default policy},
 plays uniformly at random from the legal moves until the game ends (in Hex, when one player has a winning path). 
The continuation policy has a large effect on the strength of the MCTS player.
In fact, one of the top Go engines, MoGo, uses a near-zero exploration coefficient in its child selection policy,
putting most of the control of the search in the hands of the continuation policy~\cite{gelly2007combining}.
% and the Go literature has a large body of work dedicated to improving it.
Initial research in MCTS Go players focused on using a strong continuation policy, with expert-derived heuristics to select the next move~\cite{chaslot2010adding}. 
The top Computer Go programs all have complex hand-crafted policies with rules to examine the current position and select a ``good'' move to play, possibly with a stochastic choice between several candidates. 
Only when none of the rules match do they fall back to a uniform random move. 
It was quickly found that building a stronger continuation policy does not guarantee better performance when used in the full tree search~\cite{gelly2006modification}. 
In other words, when the continuation policies are used as players, the relative strengths of the continuation policies are poor predictors
of how well the MCTS algorithms relying on these continuation policies will perform.
This surprising result may be due to stronger continuation policies under-exploring the search space.
In any case, this means that hand crafting an effective continuation policy requires benchmarking and frequent tests of the entire
MCTS system as small changes to the simulation policy can have unpredictable effects on overall performance.
This illustrates the need for automated systems for tuning the continuation policy.

@@

\subsection{Update Policy}
Once a simulation is finished, the result is used to update the search tree.
Basic MCTS systems updated the win and visit statistics only along the traversed path.
It was quickly discovered that more aggressive update strategies are often profitable.
One simple example is the All-Moves-As-First (AMAF) heuristic~\cite{brugmann1993monte}
that exploits the fact that moves played in a simulation could have been played in any order. 
Thus, a move deep in the simulation could be fictitiously treated as the ``first'' move from an intermediate position. 
In addition to updating the statistics of the nodes in the path, all children of nodes on the path corresponding to moves made later
in the simulation also have their AMAF win and visit counts updated. 
%Variations on AMAF have been tested, including Permutations AMAF, which updates a node if any permutation of the simulation contained the node's position, and Some-Moves-As-First, which attempts to restrict the number of siblings updated.

@@

% TODO
%\textbf{does MoHex do something fancier?}
% MoHex uses one local pattern - bridge preservation, and AMAF updates with RAVE (http://webdocs.cs.ualberta.ca/~hayward/hex/#mohex)

@@

@@

Hivemind is a system that learns to play abstract strategy games on a regular board. 
The system consists of three modules: the evolutionary learning process, a Monte-Carlo based UCT search tree implementation, and a ``Game Tracker interface'' used by the tree search to maintain game state information. 
The design was influenced by Fuego, although Hivemind is written in a different language (Google's Go~\cite{golang}). 
In particular, the tracker interface is very similar to Fuego's subclassing system for implementing multiple games~\cite{Fuego}. 
Hivemind's tracker interface is currently implemented for the games of Hex, Go, and Tic-Tac-Toe. 
The major contribution of Hivemind is the integration of evolutionary learning to automatically learn appropriate continuation policies.
% Hivemind is also open-source software, available at \url{http://github.com/etherealmachine/hivemind}.
Preliminary work tried learning with particle swarm optimization, but we found that evolution strategies gave
superior results.

Using evolution strategies to learn continuation policies requires that the continuation policies are encoded as vectors.
We first describe this encoding.  Section~\ref{s:tracker} describes the Game Tracker module, and Section~\ref{s:es} describes the
use of Evolutionary Strategies.

@@

\begin{figure}
	\begin{center}
	\includegraphics[width=3.0in]{graphics/local-pattern.pdf}
	\caption{A 4x4 board with the 3 patterns shown around \hblacks\ last move. Example weights are shown in the center vertex.}
	\label{fig:localpattern}
	\end{center}
\end{figure}

\begin{figure}
	\begin{center}
	\includegraphics[width=3.0in]{graphics/weight-pattern-map.pdf}
	\caption{The encoding from pattern A to a 32-bit integer. Using the weight from figure \ref{fig:localpattern}, the real value 6 would be stored at position 102 in the weight map.}
	\label{fig:encoding}
	\end{center}
\end{figure}
